---
title: "Descriptive statistic: Tabular and Graphical Presentations"
output: html_document
editor_options: 
  chunk_output_type: console
---

## Summarizing Categorical Data

### **Frequency Distribution**

A frequency distribution is a tabular summary of data showing the number (frequency) of items in each of several non overlapping classes.

**Example 2.1** Consider the following data shown in @tbl-tbl1.

|              |              |              |
|:-------------|:-------------|:-------------|
| Coke Classic | Coke Classic | Coke Classic |
| Diet Coke    | Diet Coke    | Coke Classic |
| Pepsi        | Coke Classic | Pepsi        |
| Diet Coke    | Diet Coke    | Dr. Pepper   |
| Coke Classic | Coke Classic | Coke Classic |
| Coke Classic | Sprite       | Diet Coke    |
| Dr. Pepper   | Pepsi        | Pepsi        |
| Diet Coke    | Coke Classic | Pepsi        |
| Pepsi        | Coke Classic | Pepsi        |
| Pepsi        | Coke Classic | Pepsi        |
| Coke Classic | Pepsi        | Coke Classic |
| Dr. Pepper   | Coke Classic | Dr. Pepper   |
| Sprite       | Sprite       | Pepsi        |
| Coke Classic | Dr. Pepper   | Sprite       |
| Diet Coke    | Pepsi        | Coke Classic |
| Coke Classic | Diet Coke    | Sprite       |
| Coke Classic | Pepsi        |              |

: Data from a sample of 50 soft drink purchases {#tbl-tbl1}

Now we will construct a frequency distribution by simply counting each type of soft-drink.

```{r }
#| label: tbl-tbl2
#| tbl-cap: "Frequency distribution of Soft Drink Purchases"

library(readxl)
library(tidyverse)
library(knitr)
MBA <- read_excel("StatForBandE_data.xlsx",sheet = "Sheet1",range = "A1:A51")

MBA %>% count(`Soft Drink`) %>% kable(col.names = c("Soft Drink","Frequency"),align = c("l","c"))
```

**Relative Frequency and Percent Frequency Distributions**

-   Relative Frequency $=\frac{Frequency \ \ of \ \ the \ \ class}{n}$

-   The *percent frequency* of a class is the relative frequency multiplied by 100.

### Bar Charts and Pie Charts

-   **Bar chart:** A graphical device for depicting qualitative data that have been summarized in a frequency, relative frequency, or percent frequency distribution.

-   **Pie chart:** A graphical device for presenting data summaries based on subdivision of a circle into sectors that correspond to the relative frequency for each class.

From the frequency table of soft drinks purchase, we will develop relative and percent frequency distribution (see @tbl-tbl3) and will construct a **bar-chart** and **pie-chart**.

```{r}
#| label: tbl-tbl3
#| tbl-cap: "Frequency, Relative And Percent Frequency Distributions Of Soft Drink Purchases"
#| 

MBA %>% count(`Soft Drink`) %>% mutate(RF=n/sum(n),PF=RF*100) %>%
  kable(digits = 2,col.names = c("Soft Drink","Frequency (f)","Relative Frequency(Rf)", "Percent Frequency (Pf)"),
        align = c("l","c","c","c"))
  
```

Now we construct a bar chart and pie chart.

```{r}
#| label: fig-fig1
#| fig-cap: "Bar chart of Soft drink purchases"
#| fig-align: left
#| fig-height: 5
#| fig-width: 5
#| fig-asp: 0.618


library(patchwork)

bar<-MBA %>% ggplot(aes(x=`Soft Drink`,fill=`Soft Drink`))+
  geom_bar(color="black",lwd=.7)+
  scale_y_continuous(breaks = seq(0,30,2))+
  guides(fill=FALSE)+
  theme_classic()+
  #labs(title = "BAR CHART OF SOFT DRINK PURCHASES",y="Frequency")+
  theme(axis.text=  element_text(color = "black"))
  
bar

```

```{r}
#| label: fig-fig2
#| fig-cap: "Pie chart of Soft drink purchases"
#| fig-align: left
#| fig-height: 5
#| fig-width: 5
#| fig-asp: 0.618

pi_data<-MBA %>% count(`Soft Drink`) %>% mutate(RF=n/sum(n),labels = scales::percent(RF))


pi_chart<-ggplot(pi_data, aes(x = "", y = RF, fill = `Soft Drink`)) +
  geom_col() +
  geom_label(aes(label = labels),color = c("white","white","white","white","black"),
            position = position_stack(vjust = 0.5),
            show.legend = FALSE,color="white") +
  guides(fill = guide_legend(title = "Soft Drink")) +
  scale_fill_viridis_d() +
  coord_polar(theta = "y") + 
  #labs(title = "PIE CHART OF SOFT DRINK PURCHASES")+ 
  theme_void()

pi_chart
```

### Cross-tabulation and its graphical presentation

**Cross-tabulation** (also known as crosstabs or contingency tables) is a fundamental data analysis technique used to examine the relationship between **two** or **more** **categorical variables.** For instance, consider the following cross-table between the **gender** and **fitness-level** of 30 individuals:

```{r echo=FALSE}
library(readr) # Recommended for robust CSV reading

data_string <- "ID,Age,Gender,Monthly Income (USD),Hours of Exercise/Week,Favorite Fruit,Fitness Level,Temperature Preference (°C)
1,25,Male,1520.75,2.5,Mango,Fair,23
2,32,Female,2280.50,1.2,Apple,Poor,22
3,20,Male,925.10,4.8,Banana,Good,19
4,29,Female,1980.90,3.3,Orange,Fair,22
5,40,Male,3055.45,0.7,Apple,Poor,21
6,23,Female,1425.00,5.9,Mango,Excellent,20
7,37,Male,3625.80,2.1,Banana,Fair,18
8,31,Female,2520.00,1.5,Apple,Poor,27
9,27,Male,1685.25,3.7,Orange,Good,22
10,22,Female,1190.00,6.2,Mango,Excellent,19
11,34,Male,3180.45,0.0,Banana,Poor,24
12,26,Female,1830.60,4.3,Orange,Good,19
13,39,Male,3999.00,1.4,Apple,Fair,22
14,24,Female,1335.75,5.6,Banana,Excellent,26
15,28,Male,2125.35,2.2,Mango,Fair,27
16,33,Female,2885.50,3.1,Orange,Good,22
17,21,Male,960.00,6.0,Apple,Excellent,18
18,30,Female,2190.00,1.7,Banana,Fair,27
19,36,Male,3425.90,2.3,Mango,Fair,27
20,29,Female,1920.40,4.5,Orange,Good,20
21,35,Male,3095.25,3.0,Apple,Good,23
22,22,Female,1125.00,5.1,Banana,Excellent,27
23,38,Male,4180.60,1.8,Mango,Fair,22
24,30,Female,2590.00,2.4,Orange,Fair,20
25,41,Male,3540.85,0.9,Apple,Poor,22
26,28,Female,1875.50,4.0,Banana,Good,24
27,24,Male,1275.00,6.7,Mango,Excellent,21
28,31,Female,2390.25,3.5,Orange,Good,25
29,36,Male,2965.80,2.6,Banana,Fair,22
30,21,Female,985.00,5.4,Apple,Excellent,25"

# Read the data into a data frame
df_r <- read_csv(data_string)

# View the structure of the data frame
#str(df_r)

# View the first few rows
#head(df_r)
```

```{r echo=FALSE}
#| label: tbl-cross.tab
#| tbl-cap: "Cross-Tabulation of Gender by Fitness Level"

# Load the knitr package for kable function
library(knitr)
library(dplyr) # For data manipulation if needed, though not strictly necessary for this simple table

# Create the data for the cross-tabulation
# It's good practice to define your data explicitly

data_table <- data.frame(
  Gender = c("Male", "Female", "Grand Total"),
  Excellent = c(2, 5, 7),
  Fair = c(7, 3, 10),
  Good = c(3, 5, 8),
  Poor = c(3, 2, 5),
  `Grand Total` = c(15, 15, 30),
  check.names = FALSE # This allows column names with spaces like "Grand Total"
)

# Generate the kable table
# Quarto automatically handles the rendering for different output formats

kable(data_table,
      format = "markdown", # Or "html", "latex", etc., Quarto often infers
      align = c("l", "c", "c", "c", "c", "c"), # Left-align Gender, center-align others
      )

```

We can show the information of cross-tab using (a) Clustered column chart or (b) Stacked column chart (see @fig-stacked.chart ).

```{r echo=FALSE}
#| label: fig-stacked.chart
#| fig-cap: "Visulization of relationship between two categorical variables"
#| fig-height: 6


# Reshape the data from wide to long format for ggplot2
# This makes it easier to plot 'Fitness Level' as a fill or x-axis variable
library(tidyr)

df_r$`Fitness Level`<- factor(df_r$`Fitness Level`,
                                  levels = c("Poor", "Fair", "Good", "Excellent"))

# Create the grouped bar chart
library(ggplot2)

ggplot(df_r, aes(x = Gender, fill = `Fitness Level`)) +
  geom_bar(position = position_dodge(width = 1)) +
  geom_text(
    stat = "count",
    aes(label = ..count..),
    position = position_dodge(width = 1),
    vjust = 0.2,
    size = 4
  ) +
  labs(
    title = "Cluster bar chart: Fitness Level distribution by Gender",
    x = "Gender",
    y = "Number of Individuals"
  ) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 10)) +  # Discrete y-axis
  scale_fill_discrete(name = "Fitness Level") +
  facet_wrap(~ Gender, scales = "free_x")+
  theme_minimal()+
  theme(strip.text = element_blank())->cluster.fig




ggplot(df_r, aes(x = Gender, fill = `Fitness Level`)) +
  geom_bar() +
  geom_text(
    stat = "count",
    aes(label = ..count..),
    position = position_stack(vjust = 0.5),
    size = 4
  ) +
  labs(
    title = "Stacked Bar Chart: Fitness level by Gender",
    x = "Gender",
    y = "Number of Individuals"
  ) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 10)) +
  scale_fill_discrete(name = "Fitness Level") +
  theme_minimal()->stacked.fig 

library(patchwork)
cluster.fig/stacked.fig+  plot_annotation(tag_levels = list(c('(a)', '(b)'), '1'))

```

## Summarizing Quantitative Data

### **Frequency Distribution** of quantitative data

Consider the following data.

YEAR-END AUDIT TIMES (IN DAYS): 12, 14, 19, 18, 15, 15, 18, 17, 20, 27, 22, 23, 22, 21, 33, 28, 14, 18, 16, 13,

```{r}
Audit<-c(12, 14, 19, 18, 15, 15, 18, 17,
20, 27, 22, 23, 22, 21, 33, 28,14, 18, 16, 13)
#summary(Audit)
```

To construct a frequency distribution we have to

1.  Determine the *number of non overlapping classes(k)*.
2.  Determine the *width* of each class.
3.  Determine the *class limits*.

### Frequency Distribution of Audit time data

Here, $n=20$, Smallest value=12, Largest value=33.

1.  Determine number of classes, $k$ as : $k=\sqrt n=\sqrt 20=4.47\approx5$. So $5$ is the number of *classes*.
2.  Class width $w$ as: $w=\frac{Largest-Smallest}{k}=\frac{33-12}{5}=4.2\approx 5$
3.  Class limits: Start from near *smallest* value (12) say from $10$ we have the following classes ([*exclusive method-where upper bound of the class is excluded*]{.underline}):

**\[10,15), \[15,20), \[20,25), \[25,30), and \[30,35)**

Now count the data values in corresponding classes and thus we have the *frequency distribution*. Once we have the frequency distribution then we also can produce the *relative* and *percent frequency distribution* (@tbl-tbl4 ).

```{r}
#| label: tbl-tbl4
#| tbl-cap: "Frequency, relative frequency (rf) and percent frequency (pf) distribution for the audit time data (n=20)" 

data.frame(Audit)->fd_data

fd_data %>% mutate(Audit_clas=ifelse(Audit%in%c(10:14),"[10,15)",
                              ifelse(Audit%in%c(15:19),"[15,20)",
                              ifelse(Audit%in%c(20:24),"[20,25)",                                         ifelse(Audit%in%c(25:29),"[25,30)","[30,35)")))))->fd

fd %>% count(Audit_clas) %>% mutate(rf=n/sum(n),pf=100*rf)->fd2 

fd2 %>% kable(digits = 2,col.names = c("Audit Time (days)","Frequency (f) ","Relative frequency (rf)", "Percent frequency(pf)"),align = "c")

```

### Histogram

A common graphical presentation of quantitative data is a *histogram.* This graphical summary can be prepared for data previously summarized in either a *frequency*, *relative frequency*, or *percent frequency* distribution.

```{r}
#| label: fig-fig3
#| fig-cap: "Histogram for the Audit Time data"
#| fig-align: left
#| fig-height: 5
#| fig-width: 5
#| fig-asp: 0.618

#png(filename="HISTOGRAM.png", width=600, height=600)

fd2 %>% ggplot(aes(x=Audit_clas,y=n))+geom_col(width =1,col="black",fill="steelblue")+
  scale_y_continuous(breaks = 0:8)+
  theme_classic()+
  labs(x="Audit Time (days)",y="Frequency")+
  theme(axis.text = element_text(size = 12))

#dev.off()


```

**Important Note:** A **relative frequency/ percent frequency histogram** is ideal for comparing distributions across groups of different sizes, as it displays proportions instead of raw counts, allowing fair and meaningful comparisons.

**Illustration ( see** @fig-rfhist.is.better**)**

A dataset of **Marks** (out of 100) was collected from two student groups:

-   **70 Female students**

-   **40 Male students**

The goal is to compare the distribution of marks between these two groups.

The **frequency histogram** shows how many students fall into each marks range (bin). However, since the number of female students (70) is greater than male students (40), their bars are naturally taller — even if the relative performance is similar. This makes direct comparison **unfair and misleading**.

A **relative frequency histogram** shows the **proportion** of students in each bin **within each group**. By dividing counts by the total number in the group:

-   It **normalizes** the data,

-   Allows for **fair comparisons** between groups of different sizes,

-   Highlights **true differences in distribution**, not just differences in group size.

**Example Observation**

From the RF histogram:

-   Around **40% of males** scored between 50–60.

-   Around **39% of females** scored between 50–60.

This comparison is **valid** only because the histograms show **relative frequency**, not raw counts.

```{r}
#| label: fig-rfhist.is.better
#| fig-cap: "Comparison between Frequency histogram and relative frequency histogram"
#| fig-width: 8
#| fig-height: 6

set.seed(332)  # For reproducibility

# Generate 30 male CGPAs and 50 female CGPAs
male_cgpa <- round(rnorm(40,70,10), 2)
female_cgpa <- round(rnorm(70, 70,10), 2)

# Combine into a data frame
df <- data.frame(
  StudentID = 1:110,
  Gender = c(rep("Male", 40), rep("Female", 70)),
  Marks = c(male_cgpa, female_cgpa)
)

library(psych)
#describe(df)


#head(df)

library(dplyr)
library(ggplot2)

# Step 1: Bin the Marks and calculate count per gender
df_binned <- df %>%
  mutate(bin = cut(Marks, breaks = seq(30, 100, by = 10), right = FALSE)) %>%
  group_by(Gender, bin) %>%
  summarise(count = n(), .groups = "drop")

# Step 2: Plot frequency histogram using geom_col
ggplot(df_binned, aes(x = bin, y = count, fill = Gender)) +
  geom_col(color = "black",width = 1) +
  geom_text(aes(label = count), vjust = 1) +
  facet_wrap(~Gender,labeller = labeller(Gender = c("Male" = "Male(n=40)", "Female" = "Female (n=70)"))) +
  labs(
    x = "Marks",
    y = "Number of Students / Frequency",
    subtitle = "Frequency Histogram of Marks by Gender"
  ) +
  #annotate()
  theme_minimal() +
  theme(legend.position = "none",
        axis.text = element_text(size = 10),
        strip.text = element_text(size = 11, 
                                  face = "bold"),
          plot.background = element_rect(color = "black", size = 1, fill = "beige"))->freqhist.Marks

  


# Step 1: Create bin variable and calculate counts per bin and gender
df_binned <- df %>%
  mutate(bin = cut(Marks, breaks = seq(30, 100, by = 10), right = FALSE)) %>%
  group_by(Gender, bin) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(Gender) %>%
  mutate(rf = count / sum(count))  # relative frequency per gender

# Step 2: Plot using geom_col
ggplot(df_binned, aes(x = bin, y = rf, fill = Gender)) +
  geom_col(color = "black",width = 1) +
  geom_text(aes(label = round(rf, 2)), vjust = 1) +
  facet_wrap(~Gender,labeller = labeller(Gender = c("Male" = "Male(n=40)", "Female" = "Female (n=70)"))) +
  labs(
    x = "Marks",
    y = "Relative Frequency",
    subtitle = "Relative Frequency Histogram of Marks by Gender"
  ) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text = element_text(size = 10),
        strip.text = element_text(size = 11, 
                                  face = "bold"),
         plot.background = element_rect(color = "black", size = 1, fill = "beige"))->rfreqhist.Marks


library(patchwork)

rf.better<-freqhist.Marks/rfreqhist.Marks

rf.better

###




```

### HISTOGRAM and shape of the distribution

See @fig-hist.skewness .

```{r}
#| label: fig-hist.skewness
#| fig-cap: "Histograms Showing Differing Levels of Skewness"

# Load required libraries
library(ggplot2)
library(dplyr)

set.seed(123)

# Generate data with different skewness
df <- data.frame(
  value = c(
    rbeta(500, 2, 5),    # Moderately right skewed
    rbeta(500, 5, 2),    # Moderately left skewed
    rnorm(500, 0.5, .2),# Symmetric
    rbeta(500, 1, 5)     # Highly right skewed
  ),
  panel = factor(rep(c("B: Moderately Skewed Right",
                       "A: Moderately Skewed Left",
                       "C: Symmetric",
                       "D: Highly Skewed Right"),
                     each = 500),
                 levels = c("A: Moderately Skewed Left",
                            "B: Moderately Skewed Right",
                            "C: Symmetric",
                            "D: Highly Skewed Right"))
)

# Plot
ggplot(df, aes(x = value)) +
  geom_histogram(aes(y = after_stat(density)), bins = 10, fill = "#58A0C8", color = "black") +
  facet_wrap(~panel, ncol = 2) +
  labs(
    title = " ",
    y = NULL,
    x = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold", hjust = 0.5),
    panel.grid.minor = element_blank()
  )


```

### Cumulative Distributions and Ogive

A variation of the frequency distribution that provides another tabular summary of quantitative data is the *cumulative frequency* distribution. @tbl-tab.crf shows the **cumulative relative frequency** of `Audit Time` data.

```{r}
#| label: tbl-tab.crf
#| tbl-cap: "Frequency, relative frequency and Cumulative relative frequency distribution of Audit Time Data"
#| fig-width: 5
#| fig-height: 4

fd %>% count(Audit_clas) %>% mutate(rf=n/sum(n),crf=cumsum(rf)) %>%
  kable(digits = 2,col.names = c("Audit Time (days)","Frequency (f) ","Relative frequency (rf)", "Cumulative relative frequency(crf)"),align = "c")
  
```

### Ogive

Another way of presenting this information is the ogive, which is a graphical representation of the cumulative relative frequencies. @fig-ogive is the drawn ogive for the cumulative relative frequency for `Audit time` data.\

```{r}
#| label: fig-ogive
#| fig-cap: "Ogive for Audit Time data"


# Load library
library(ggplot2)

# Data frame
data <- data.frame(
  AuditTime = c(10,15, 20, 25, 30, 35),
  CRF = c(0,0.20, 0.60, 0.85, 0.95, 1.00)
)

# Plot
ggplot(data, aes(x = AuditTime, y = CRF)) +
  geom_line(color = "blue") +
  geom_point(size = 3, color = "blue") +
  geom_text(aes(label = sprintf("%.2f", CRF)), vjust = 2, 
            hjust=1.2,size = 4) +
  labs(
    title = " ",
    x = "Audit Time (days)",
    y = "Cumulative relative frequency"
  ) +
  scale_y_continuous(limits = c(0, 1.1),breaks = seq(0, 1, by = 0.1),
                     expand = c(0, 0)) +
  scale_x_continuous(limits = c(10,40),
                     breaks = seq(10, 36, by = 5),expand = c(0, 0))+
  theme_classic()+
  theme(axis.text = element_text(size = 12),
        axis.title = element_text(size = 14))

```

For instance, from both @tbl-tab.crf and @fig-ogive we can say (estimate) that 60% of audits took **less than 20 days** . Similarly, 95% of audits took **less than 30 days** and so on.

### The Stem-and-Leaf Display

The techniques of **exploratory data analysis** consist of simple arithmetic and easy-to-draw graphs that can be used to summarize data quickly. One technique—referred to as a **stem-and-leaf display**—can be used to show both the rank order and shape of a data set simultaneously [@anderson2011].

**Steps to Construct a Stem-and-Leaf Diagram**

\(1\) Divide each number into two parts: a **stem**, consisting of one or more of the leading digits, and a **leaf**, consisting of the remaining digit.

\(2\) List the stem values in a vertical column.

\(3\) Record the leaf for each observation beside its stem.

\(4\) Write the units for stems and leaves on the display.\

**Example 2.2** Here are the number of questions answered correctly on an aptitude test given to 50 individuals recently interviewed for a position at Haskens Manufacturing.

::: panel-tabset
## Data

112, 72, 69, 97, 107,73, 92, 76, 86, 73, 126, 128, 118, 127, 124,82, 104, 132, 134, 83, 92, 108, 96, 100, 92,115, 76, 91, 102, 81, 95, 141, 81, 80, 106,84, 119, 113, 98, 75, 68, 98, 115, 106, 95,100, 85, 94, 106, 119

## Stem-and-leaf display

```{r}

stemleaf=c(112,72,69,97,107,73,92,76,86,73,126,128,118,127,124,82,104,132,134,83,92,108,96,100,92,115,76,91,102,81,95,141,81,80,106,84,119,113,98,75,68,98,115,106,95,100,85,94,106,119)

#summary(stemleaf)

stem(stemleaf)
```
:::

**Exception**

In some data sets, providing more classes or stems may be desirable. One way to do this would be to modify the original stems as follows: For example, divide stem 5 into two new stems, 5L and 5U. Stem 5L has leaves 0, 1, 2, 3, and 4, and stem 5U has leaves 5, 6, 7, 8, and 9. This will double the number of original stems. However, there may be various type of data in practical situations. So, we have to figure out the suitable stem-and-leaf plot.

**Example 2.3:** Construct a stem-and-leaf plot from the following data:

88.5, 98.8, 89.6, 92.2, 92.7, 88.4, 87.5, 90.9, 94.7, 88.3, 90.4, 83.4, 87.9, 92.6, 87.8, 89.9, 84.3, 90.4, 91.6, 91.0

::: panel-tabset
## Solution-I

```{r}
sl<-c(88.5, 98.8, 89.6, 92.2, 92.7, 88.4, 87.5, 90.9,94.7, 88.3, 90.4, 83.4, 87.9, 92.6, 87.8, 89.9,84.3, 90.4, 91.6, 91.0)

stem(sl)
```

## Solution-II

```{r}
stem(sl,scale = 2)
```
:::

**Example 2.4 (Another example):** Construct a stem-and-leaf plot from the following data: 7,8,2,1,8,3,5,7,1,2,2,5,8,5,5,7,8,7,5,3

[Solution:]{.underline}

```{r}
singldigit=c(7,8,2,1,8,3,5,7,1,2,2,5,8,5,5,7,8,7,5,3)
stem(singldigit,2)
```

## Exercises

**2.1** A doctor’s office staff studied the waiting times for patients who arrive at the office with a request for emergency service. The following data with waiting times in minutes were collected over a one-month period.

> 2, 5, 10, 12, 4, 4, 5, 17, 11, 8, 9, 8, 12, 21, 6, 8, 7, 13, 18, 3

Use class interval/width of 5 in the following (start your class limit from 0):

a.  Show the frequency distribution.
b.  Show the relative frequency distribution.
c.  Show the cumulative frequency distribution.
d.  Show the cumulative relative frequency distribution.
e.  What proportion of patients needing emergency service wait less than 10 minutes or less?

<!-- -->

**2.2** A shortage of candidates has required school districts to pay higher salaries and offer extras to attract and retain school district superintendents. The following data show the annual base salary (\$1000s) for superintendents in 20 districts in the greater Rochester, New York, area (The Rochester Democrat and Chronicle, February 10, 2008).

> 187, 184, 174, 185, 175, 172, 202, 197, 165, 208, 215, 164, 162, 172, 182, 156, 172, 175, 170, 183

Use appropriate number classes/ class width in the following.

a.  Show the frequency distribution.
b.  Show the percent frequency distribution.
c.  Show the cumulative percent frequency distribution.
d.  Develop a histogram for the annual base salary.
e.  Do the data appear to be skewed? Explain.
f.  Which salary range belongs to the highest percentage of superintendents ?

::: panel-tabset
## Data

> 187, 184, 174, 185, 175, 172, 202, 197, 165, 208, 215, 164, 162, 172, 182, 156, 172, 175, 170, 183

## Histogram

```{r}
salary=c(187, 184, 174, 185, 175, 172, 202, 197, 165, 208, 215, 164, 162, 172, 182, 156, 172, 175, 170, 183)

#sort(salary)

#summary(salary)

hist(salary,
     breaks = seq(155,225,10),
     right = FALSE,
     xaxt = 'n',labels = TRUE,ylim = c(0,7))
axis(1, at = seq(from = 155, to = 225, by = 10))
#hist(salary,xlim = c(155, 220),right = FALSE)
```
:::

**2.3** NRF/BIG research provided results of a consumer holiday spending survey (USA Today, December 20, 2005). The following data provide the dollar amount of holiday spending for a sample of 25 consumers.

> 1200, 850, 740, 590, 340, 450, 890, 260, 610, 350, 1780, 180, 850,2050, 770, 800, 1090, 510, 520, 220, 1450, 280, 1120, 200 350

a.  What is the lowest holiday spending? The highest?
b.  Use a class width of \$250 to prepare a frequency distribution and a percent frequency distribution for the data.
c.  Prepare a histogram and comment on the shape of the distribution.
d.  What observations can you make about holiday spending?

::: panel-tabset
## Data

> 1200, 850, 740, 590, 340, 450, 890, 260, 610, 350, 1780, 180, 850,2050, 770, 800, 1090, 510, 520, 220, 1450, 280, 1120, 200, 350

## Histogram

```{r}
spending=c(1200, 850, 740, 590, 340, 450, 890, 260, 610, 350, 1780, 180, 850,2050, 770, 800, 1090, 510, 520, 220, 1450, 280, 1120, 200, 350)

#summary(spending)

hist(spending,breaks = seq(180,2180,250),xaxt="n",main = "Frequency histogram of holiday spending (USD)",xlab = "Holiday spending",labels = TRUE,ylim = c(0,9))
axis(1, at = seq(180,2180,250))

```
:::

**2.4** Construct a stem-and-leaf display for the following data.

> 70, 72, 75, 64, 58, 83, 80, 82, 76, 75, 68, 65, 57, 78, 85, 72

**2.5** Construct a stem-and-leaf display for the following data.

> 11.3, 9.6, 10.4, 7.5, 8.3, 10.5, 10.0, 9.3, 8.1, 7.7, 7.5, 8.4, 6.3, 8.8

**2.6** A psychologist developed a new test of adult intelligence. The test was administered to 20 individuals, and the following data were obtained.

> 114, 99, 131, 124, 117, 102, 106, 127, 119, 115,98, 104, 144, 151, 132, 106, 125, 122, 118, 118

**Construct** a stem-and-leaf display for the data.

## Line Chart

## Scatter diagram

## Case Study: Lifestyle Indicators and Preferences

You’re given data from a cross-sectional survey of 30 individuals. Variables include Age, Gender, Monthly Income, Hours of Exercise/Week , Fitness Level, Favorite Fruit and Temperature preference. This data-set enables practice with data visualization, scale classification, and relationship analysis using tools like bar charts, scatter plots, and histograms.

| **ID** | **Age** | **Gender** | **Monthly Income (USD)** | **Hours of Exercise/Week** | **Favorite Fruit** | **Fitness Level** | **Temperature Preference (°C)** |
|:--:|:--:|:--:|:--:|:--:|:--:|:--:|:--:|
| 1 | 25 | Male | 1520.75 | 2.5 | Mango | Fair | 23 |
| 2 | 32 | Female | 2280.50 | 1.2 | Apple | Poor | 22 |
| 3 | 20 | Male | 925.10 | 4.8 | Banana | Good | 19 |
| 4 | 29 | Female | 1980.90 | 3.3 | Orange | Fair | 22 |
| 5 | 40 | Male | 3055.45 | 0.7 | Apple | Poor | 21 |
| 6 | 23 | Female | 1425.00 | 5.9 | Mango | Excellent | 20 |
| 7 | 37 | Male | 3625.80 | 2.1 | Banana | Fair | 18 |
| 8 | 31 | Female | 2520.00 | 1.5 | Apple | Poor | 27 |
| 9 | 27 | Male | 1685.25 | 3.7 | Orange | Good | 22 |
| 10 | 22 | Female | 1190.00 | 6.2 | Mango | Excellent | 19 |
| 11 | 34 | Male | 3180.45 | 0.0 | Banana | Poor | 24 |
| 12 | 26 | Female | 1830.60 | 4.3 | Orange | Good | 19 |
| 13 | 39 | Male | 3999.00 | 1.4 | Apple | Fair | 22 |
| 14 | 24 | Female | 1335.75 | 5.6 | Banana | Excellent | 26 |
| 15 | 28 | Male | 2125.35 | 2.2 | Mango | Fair | 27 |
| 16 | 33 | Female | 2885.50 | 3.1 | Orange | Good | 22 |
| 17 | 21 | Male | 960.00 | 6.0 | Apple | Excellent | 18 |
| 18 | 30 | Female | 2190.00 | 1.7 | Banana | Fair | 27 |
| 19 | 36 | Male | 3425.90 | 2.3 | Mango | Fair | 27 |
| 20 | 29 | Female | 1920.40 | 4.5 | Orange | Good | 20 |
| 21 | 35 | Male | 3095.25 | 3.0 | Apple | Good | 23 |
| 22 | 22 | Female | 1125.00 | 5.1 | Banana | Excellent | 27 |
| 23 | 38 | Male | 4180.60 | 1.8 | Mango | Fair | 22 |
| 24 | 30 | Female | 2590.00 | 2.4 | Orange | Fair | 20 |
| 25 | 41 | Male | 3540.85 | 0.9 | Apple | Poor | 22 |
| 26 | 28 | Female | 1875.50 | 4.0 | Banana | Good | 24 |
| 27 | 24 | Male | 1275.00 | 6.7 | Mango | Excellent | 21 |
| 28 | 31 | Female | 2390.25 | 3.5 | Orange | Good | 25 |
| 29 | 36 | Male | 2965.80 | 2.6 | Banana | Fair | 22 |
| 30 | 21 | Female | 985.00 | 5.4 | Apple | Excellent | 25 |

**Tasks**

1)  **Identify** the *scale of measurement* of each variable.

2)  **Construct** suitable graph/chart for variables like *gender*, *fitness level* etc.

3)  **Construct** relative frequency histogram for variables like age, Monthly Income (USD), etc. **Describe** what you have learned.

4)  **Plot** Hours of Exercise/Week vs Temperature Preference (°C). What is your conclusion.

5)  **Draw** the scatter plot of Monthly Income vs Hours of Exercise/Week. Is there any relation?

6)  **Cross-tabulation**

> a)  Make a cross-tabulation between Gender and Favorite Fruit. Show the result in a stacked bar-chart.

> b)  Make a cross-tabulation between Favorite Fruit and Fitness Level. Show the result in a stacked bar-chart.
