---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Introduction to estimation

In previous chapter we discuss the sampling properties of the sample mean and variance. In this chapter we discuss about the **parameter estimation**.It falls under the branch of **Statistical Inference.** The process of estimation involves determining the approximate value of a population parameter on the basis of sample data. There are two types **parameter estimation-(i) point estimation** and **(ii) interval estimation.**

## Point Estimation

-   To estimate the value of a **population parameter,** we compute a corresponding characteristic of the sample, referred to as a **sample statistic**.

-   By making the preceding computations, we perform the statistical procedure called **point estimation**. For instance, we refer to the sample mean $\bar x$ as the **point estimator** of the population mean $\mu$.

-   The numerical value obtained for , is called the **point estimate**.

+-------------------------------+-------------+---------------------------------------------------------------------------------------+
| Population parameter          | Symbol      | Point estimator                                                                       |
+===============================+:===========:+:=====================================================================================:+
| Population mean               | $\mu$       | Sample mean,                                                                          |
|                               |             |                                                                                       |
|                               |             | $\bar x=\frac{\sum x}{n}$                                                             |
+-------------------------------+-------------+---------------------------------------------------------------------------------------+
| Population standard deviation | $\sigma$    | Sample standard deviation,                                                            |
|                               |             |                                                                                       |
|                               |             | $s=\sqrt{\frac{\sum(x-\bar x)^2}{n-1}}=\sqrt {\frac{\sum x^2 -n\cdot \bar x^2}{n-1}}$ |
+-------------------------------+-------------+---------------------------------------------------------------------------------------+
| Population proportion         | $p$         | Sample proportion,                                                                    |
|                               |             |                                                                                       |
|                               |             | $\hat p=\frac{\# \ \ of \ \  outcomes\ \ of  \ \ interest }{n}$                       |
+-------------------------------+-------------+---------------------------------------------------------------------------------------+

: Some common population parameters and their estimators {#tbl-tab9.1}

### Properties of Point Estimators

Suppose

$\theta$ be the population parameter of interest

$\hat \theta$ be the sample statistic or point estimator of $\theta$

A "good" estimator has some desirable properties.

::: callout-note
## Unbiasedness

A sample statistic $\hat \theta$ is said to be unbiased estimator of the population parameter $\theta$ if

$$
E(\hat\theta)=\theta
$$
:::

**Problem 9.1** Show that the function of sample mean $\bar X$ is the unbiased estimator of population mean $\mu$.

[Solution:]{.underline}

Here $X$ is the variable of interest and let $X_1,X_2,...,X_n$ is a sequence of random sample provided $E(X_i)=\mu$. The sample mean is $\bar X=\frac{\sum_{i=1}^n X_i}{n}$.

Now

$$
E(\bar X)=E\left( \frac{\sum_{i=1}^nX_i }{n}\right)=\frac{1}{n}E\left( \sum_{i=1}^n X_i\right)
$$

$$
=\frac{1}{n}\sum_{i=1}^n E(X_i)=\frac{1}{n}\sum_{i=1}^n \mu=\frac{1}{n}\cdot n\mu=\mu
$$

$$
\therefore E(\bar X)=\mu
$$

So, $\bar X$ is an unbiased estimator of $\mu$.

**Problem 9.2** Let, $X_1, X_2,...,X_n$ be a random sample from a population with mean $\mu$ and variance $\sigma^2$. That is $X_i$ 's are independent and each $X_i$ has $E(X_i)=\mu$ and $Var(X_i)=\sigma^2$.

Show that the function of sample variance $S^2=\frac{\sum_{i=1}^n(X_i-\bar X)^2}{n-1}$ is an unbiased estimator of population variance $\sigma^2$.

[Solution:]{.underline} See @newbold2013, page 283, or we can proof it as follows:

We know $E(\bar X)=\mu_{\bar X}=\mu$ .

So,

$Var(X)$ , $\sigma^2=E(X^2)-\mu^2$

Or, $E(X^2)=\mu^2+\sigma^2$

and

$Var(\bar X)$, $\sigma^2_{\bar X}=E(\bar X^2)-\mu_{\bar X}^2$

Or, $E(\bar X^2)=\mu_{\bar X}^2+\sigma^2_{\bar X}=\mu^2+\frac{\sigma^2}{n}$

Now,

$E(S^2)=E\left[ \frac{\sum_{i=1}^n (X_i-\bar X)^2}{n-1}\right]=\frac{1}{n-1}E \sum_{i=1}^n (X_i-\bar X)^2=\frac{1}{n-1}E\sum_{i=1}^n(X_i^2+\bar X^2-2 X_i \bar X)$

$=\frac{1}{n-1} E\left ( \sum_{i=1}^n X_i^2-n\bar X^2 \right)=\frac{1}{n-1} \left [\sum_{i=1}^n E(X_i^2)-n E(\bar X^2) \right]$\

$=\frac{1}{n-1} \left[ \sum_{i=1}^n (\mu^2+\sigma^2)-n(\mu^2+\frac{\sigma^2}{n}) \right]=\frac{1}{n-1} \left[ n\mu^2+n \sigma^2-n\mu^2-\sigma^2 \right]$

$=\frac{1}{n-1}(n-1) \sigma^2=\sigma^2$

$\therefore E(S^2)=\sigma^2$.

Hence $S^2$ is an unbiased estimator of $\sigma^2$.

**Important NOTE:** This proof only valid when the actual sample size, $n$, is a small proportion of the population size $N$ or for small population size samples are taken with replacement (which me).
\

::: callout-note
## Efficiency

Suppose $\hat {\theta_1}$ and $\hat {\theta_2}$ be two unbiased estimators of population parameter $\theta$.

$\hat {\theta_1}$ is said to be more efficient than $\hat {\theta_1}$ if $Var(\hat {\theta_2})<Var(\hat {\theta_2})$.
:::

::: callout-note
## Consistency

A point estimator $\hat {\theta}$ is sad to be consistent estimator of the parameter $\theta$ if the variance of the estimator becomes smaller as sample size increases.

In other word,

$$
\lim_{n\rightarrow\infty} Var(\hat {\theta})=0
$$\
:::

**Problem 9.3** [@anderson2020] A simple random sample of 30 managers and the corresponding data on annual salary and management training program participation are as shown in @tbl-tab9.2

| Annual Salary (\$000) | Management Training Program |
|:---------------------:|:---------------------------:|
|         49.09         |             Yes             |
|         53.26         |             Yes             |
|         49.64         |             Yes             |
|         49.89         |             Yes             |
|         47.62         |             No              |
|         45.92         |             Yes             |
|         49.09         |             Yes             |
|         51.40         |             Yes             |
|         50.96         |             Yes             |
|         45.11         |             Yes             |
|         45.92         |             No              |
|         57.27         |             Yes             |
|         55.69         |             No              |
|         51.56         |             No              |
|         56.19         |             No              |
|         51.77         |             Yes             |
|         52.54         |             No              |
|         44.98         |             Yes             |
|         51.93         |             Yes             |
|         52.97         |             Yes             |
|         45.12         |             Yes             |
|         51.75         |             Yes             |
|         54.39         |             No              |
|         50.16         |             No              |
|         52.97         |             Yes             |
|         50.24         |             Yes             |
|         52.79         |             Yes             |
|         50.98         |             No              |
|         55.86         |             Yes             |
|         57.31         |             No              |

: Annual Salary and Training Program Status for a Simple Random Sample of 30 EAI Managers {#tbl-tab9.2}

a\) **Compute** *sample mean* and *standard deviation* of annual salary (\$) of a random sample of 30 EAI managers.

[Solution:]{.underline}

Let $\mu$ be the population mean of annual salary of all EAI managers.

If $X$ is the annual salary in '000 USD, then the to estimate $\mu$ we use **sample mean** $\bar x$ as follows:

$$
\bar x=\frac{\sum_{i=1}^n x_i}{n}=\frac{49.09+53.26+...+57.31}{30}\approx51.1457 
$$

So the **sample mean** is $\$ 51145.7$.

Similarly let $\sigma$ be the population standard deviation of annual salary of all EAI managers.

The estimate of $\sigma$ is the **sample standard deviation** $s$ as follows:

$$
s=\sqrt \frac{\sum_{i=1}^n x_i^2 -n\cdot (\bar x)^2}{n-1}
$$

$$
=\sqrt \frac{(49.09^2+53.26^2+...+57.31^2)-30\cdot (51.1457)^2}{30-1}
$$

$$
\approx 3.5408
$$

So the **sample standard deviation** is $\$ 3540.8$.

b\) Also, **estimate** the *proportion* of managers in the population who completed the management training program.

[Solution:]{.underline} Here, $n=30$

Let, $p$ be the population proportion of managers who completed the training

The estimate of $p$ is:

$$
\hat p=\frac {\# \ \ of\ \ yes}{n}=\frac{20}{30}=0.6667\approx 66.67\%
$$

**Problem 9.4** [@anderson2020]Many drugs used to treat cancer are expensive. Business Week reported on the cost per treatment of Herceptin, a drug used to treat breast cancer (Business Week, January 30, 2006). Typical treatment costs (in dollars) for Herceptin are provided by a simple random sample of 10 patients.

4376 ,5578, 2717, 4920, 4495, 4798, 6446, 4119, 4237, 3814

a\) **Develop** a point estimate of the mean cost per treatment with Herceptin.

b\) **Develop** a point estimate of the standard deviation of the cost per treatment with Herceptin.

## Interval estimation

Instead of estimating a population parameter by a single value (point estimator) it is more reasonable to estimate with an **interval** with some confidence (probability) that our **parameter** value will be in the **interval.**

::: callout-note
## Interval Estimator

An **interval estimator** is a rule for determining (based on sample information) an interval that is likely to include the parameter. The general form of an interval estimate is as follows:

$$
Point\ \ estimate \pm margin \ \ of \ \ error
$$
:::

Due to sampling variability, **interval estimator** is also random.

### Interval estimate of a population mean: $\sigma$ known

The $(1-\alpha)100\%$ confidence interval for $\mu$ is :

$$
\bar x \pm z_{\alpha/2} \frac{\sigma}{\sqrt n}
$$ {#eq-9.1}

Or,

$$
\bar x-z_{\alpha/2}\frac {\sigma}{\sqrt n}, \bar x+z_{\alpha/2}\frac {\sigma}{\sqrt n}
$$

We can express this confidence interval in a probabilistic way:

$$
P\left( \bar x-z_{\alpha/2}\frac{\sigma}{\sqrt n}<\mu<\bar x+z_{\alpha/2}\frac{\sigma}{\sqrt n}  \right)=1-\alpha
$$

**NOTE:**

1\) Here, $z_{\alpha/2}$ is the $z$ value providing an area of $\alpha/2$ in the upper tail of the standard normal distribution that is $P(Z>z_{\alpha/2})=\alpha/2$.

2\) $z_{\alpha/2} \cdot \frac{\sigma}{\sqrt n}$ is often called **margin of error (ME)**.

### **Interpretation of confidence interval**

The probabilistic equation of confidence interval says that, if we repeatedly construct confidence intervals in this manner, we will expect $(1-\alpha)100\%$ of them contain $\mu$.

### Understanding confidence interval through Simulation

Suppose $X\sim N(50,5^2)$ . Now consider a population data of size $N=10000$ and the histogram of $X$ is:

```{r}
set.seed(36)
X<-rnorm(10000,50,5)

hist(X,freq = F,sub= "Population size, N=10000")
lines(density(X),lwd=2)
```

Now we draw a random sample of size $n=50$ from this population and construct a 95% confidence interval (CI) for $\mu$. The CI may or may not include the $\mu=50$ !!!

```{r}
set.seed(36)
mu=50;sigma=5

## Constructing (1-alpha)*100% CI

alpha=0.05 

con.coef=1-alpha # confidence level

z=round(abs(qnorm(alpha/2)),2)# z=1.96

n=50 # sample size

s.e<-sigma/sqrt(n)

sampl_1<-sample(X,n)
cat("Sample data :", sampl_1)

cat("Sample mean:",round(mean(sampl_1),2))

ci_1<-c(lower=mean(sampl_1)-z*s.e,upper=mean(sampl_1)+z*s.e)
#ci_1[1]
cat("95%  CI:","\n", "[Lower ,Upper]","\n", "[",round(ci_1[1],2),",",round(ci_1[2],2),"]")
```

Luckily our 95% CI contains the true population mean $\mu=50$ 😊.

Lets simulate 100 samples each of size $n=50$ and construct all 95% CIs.

```{r}
#| label: fig-conf_sim
#| fig-cap: "Simulation of 95% confidence intervals for $\\mu$"

library(tidyverse)

# Suppose, X~N(50,5^2); so
#cat("mu=",50,",", "sigma=",5)



# Let simulate 100 samples each of size n=50

sampl=0

B=100 # number of samples we have drawn from population X

sampl<-(replicate(B,sample(X,n,replace = FALSE)))

sample.means<-colMeans(sampl)

#class(sample.means)

sample.means<-as.data.frame(sample.means)
#class(sample.means)

sample.means%>%rename(x_bar=sample.means)->sample.means

ci<-sample.means%>%mutate(ll=x_bar-z*s.e,ul=x_bar+z*s.e)

ci%>%mutate(id=1:100)%>%select(id,x_bar,ll,ul)->ci

ci%>%mutate(Capture=ifelse(50>ll & 50<ul,"1","0"))->ci_95

#ci%>%head()

# https://statisticsglobe.com/draw-plot-with-confidence-intervals-in-r

colorset = c('0'='red','1'='black')


labels<- expression("Population mean,"~mu == 50)

ggplot(ci_95, aes(id, x_bar)) +
  geom_point() +
  geom_errorbar(aes(ymin = ll, ymax = ul,color = Capture))+
  geom_hline(yintercept = 50, linetype = "dashed", color = "blue")+
  scale_color_manual(values = colorset)+
  ylim(45,55)+
  scale_x_continuous(breaks = seq(1,100,5),limits=c(0, 101))+
  #annotate("text",label=paste("Population mean,mu=",mu),x=90,y=54)+
  annotate("text",x=90,y=53.5,label=as.character(labels),parse=TRUE)+
    labs(title =paste(con.coef*100, "% Confidence Intervals, n =", n),
       x="Sample ID")+
  coord_flip()+
  theme_bw()
```

We can see that out of 100 CIs , 95 of them contain true population mean $\mu=50$ and the rest 5 do not.

| $1-\alpha$ | $\alpha$ | $z_{\alpha/2}$ |
|:----------:|:--------:|:--------------:|
|    0.90    |   0.10   |     1.645      |
|    0.95    |   0.05   |      1.96      |
|    0.98    |   0.02   |      2.33      |
|    0.99    |   0.01   |     2.575      |

: Four Commonly Used Confidence Levels and $z_{\alpha/2}$ {#tbl-tab9.3}

### Interval estimate of a population mean: $\sigma$ unknown

The $(1-\alpha)100\%$ confidence interval for $\mu$ is :

$$
\bar x \pm t_{\alpha/2} \frac{s}{\sqrt n}
$$ {#eq-9.2}

Or,

$$
\bar x-t_{\alpha/2}\frac {s}{\sqrt n}, \bar x+t_{\alpha/2}\frac {s}{\sqrt n}
$$

We can express this confidence interval in a probabilistic way:

$$
P\left( \bar x-t_{\alpha/2}\frac{s}{\sqrt n}<\mu<\bar x+t_{\alpha/2}\frac{s}{\sqrt n}  \right)=1-\alpha
$$

Here, $t_{\alpha/2}$ is the $t$ value providing an area of $\alpha/2$ in the upper tail of the $t$ distribution with $(n-1)$ degrees of freedom that is $P(T>t_{\alpha/2,n-1})=\alpha/2$.

**Problem 9.5** [@keller2014, page 341] In a survey conducted to determine, among other things, the cost of vacations, 64 individuals were randomly sampled. Each person was asked to compute the cost of her or his most recent vacation and the sample mean was \$1810.16. Assuming that the standard deviation is \$400, **estimate** with 95% confidence interval for the average cost of all vacations.

***Solution:** Let,* $X$ be the vacation cost in \$. Also let $\mu$ be the mean of all vacations.

Here sample size is, $n=64$ with the sample mean , $\bar x =\$ 1810.16$.

The assumed (population) standard deviation, $\sigma=\$400$. Since the sample size is sufficiently large so according to CLT, $\bar X\sim N(\mu, \sigma_{\bar X}^2) \ \ approximately$ .

Where $\sigma_{\bar X}=\frac{\sigma}{\sqrt n}=\frac{400}{\sqrt 64}=50$

For 99% confidence interval (CI), $z_{\alpha/2}=2.575$.

So the 99% CI for the $\mu$ is:

$$
\bar{x} \pm z_{\alpha/2} \times \frac{\sigma}{\sqrt{n}}
$$

$$
Or,1810.16 \pm 2.575 \times 50
$$ $$
Or, 1810.16 \pm 128.75
$$

$$
[1681.41, 1938.91]
$$

$\therefore P(1681.41<\mu<1938.91)=99\%$.

**Problem 9.6** [@keller2014, page 340] It is known that the amount of time needed to change the oil on a car is normally distributed with a standard deviation of 5 minutes. The amount of time to complete a random sample of 10 oil changes was recorded and listed here. **Compute** the 99% confidence interval estimate of the mean of the population.

11, 10, 16, 15 ,18, 12 ,25,20, 18 ,24

**Problem 9.7** [@newbold2013, page 302] How much do students pay, on the average, for textbooks during the first semester of college? From a random sample of 400 students the mean cost was found to be \$357.75, and the sample standard deviation was \$37.89. Assuming that the population is normally distributed, find the 95% confidence interval for the population mean.\

**Problem 9.8** [@newbold2013, page 302]Twenty people in one large metropolitan area were asked to record the time (in minutes) that it takes them to drive to work. These times were as follows:

30, 42, 35, 40, 45, 22, 32, 15, 41, 45, 28, 32, 45, 27, 47, 50, 30, 25, 46, 25

Assuming that the population is normally distributed find the 99% confidence interval for the population mean of time it takes to drive to work.

### Interval estimation for a population proportion : Large sample

From previous chapter we know if $np$ and $np(1-p)$ is equal or greater than $5$ then the sample proportion $\hat p$ will approximately follow **normal distribution** with **mean** $p$ and **variance** $\frac{p(1-p)}{n}$.

Mathematically,

$$
Z=\frac{\hat p-p}{\sqrt {\frac{p(1-p)}{n}}}\approx follows \ \ N(0,1)
$$

Since $p$ is unknown we estimate $var(\hat p)$ as $\frac{\hat p(1-\hat p)}{n}$.

**Confidence Interval for Population Proportion,** $p$ **(Large Samples)**

$$
\hat p\pm z_{\alpha/2} \sqrt {\frac{\hat p(1-\hat p)}{n}}
$$ {#eq-9.3}\
which is valid provided that $n\hat p$ and $n(1 - \hat p)$ are greater than 5.

**Problem 9.9** [@newbold2013, page 305] In a random sample of 95 manufacturing firms, 67 indicated that their company attained ISO certification within the last two years. Find a 99% confidence interval for the population proportion of companies that have been certified within the last 2 years.

**Problem 9.10** [@newbold2013, page 305] From a random sample of 400 registered voters in one city, 320 indicated that they would vote in favor of a proposed policy in an upcoming election. Find a 98% confidence interval for the population proportion in favor of this policy.

## Sample size determination: Large Population

### Sample Size for Population Mean

If $E$ be the *margin of error **(E)**,* $\sigma$ be the population standard deviation (assume known) then the required **sample size (n)** to estimate a population mean $\mu$ with $(1-\alpha)\times 100\%$ confidence level is:

$$
n=\left( \frac{z_{\alpha/2} \sigma}{E} \right)^2
$$ {#eq-9.4}

::: callout-note
## Sample size determination: Derivation

From the interval estimation regarding $\mu$ we know

$$
P(\bar x-z_{\alpha/2} \frac{\sigma}{\sqrt n}<\mu<\bar x+z_{\alpha/2} \frac{\sigma}{\sqrt n})=1-\alpha
$$

That is,

$$
P(\bar x-E<\mu<\bar x+E)=1-\alpha
$$

Hence we can write,

$$
E=z_{\alpha/2} \frac{\sigma}{\sqrt n}
$$

Solving for $n$ we have

$$
n=\left( \frac{z_{\alpha/2} * \sigma}{E} \right)^2
$$
:::

**Problem 9.11 [@keller2014]**

a\) Determine the sample size required to estimate a population mean to within 10 units given that the population standard deviation is 50. A confidence level of 90% is judged to be appropriate.

b\) Repeat part (a) changing the standard deviation to 100.

c\) Re-do part (a) using a 95% confidence level.

d\) Repeat part (a) wherein we wish to estimate the population mean to within 20 units.\

**Problem 9.12 [@keller2014]** The operations manager of a large production plant would like to estimate the average amount of time workers take to assemble a new electronic component. After observing a number of workers assembling similar devices, she guesses that the standard deviation is 6 minutes. How large a sample of workers should she take if she wishes to estimate the mean assembly time to within 20 seconds? Assume that the confidence level is to be 99%.

### Sample Size for Population Proportion

If $E$ be the *margin of error **(E)**,* $p*$ be the assumed population proportion or planning value of $\hat p$. Then the required **sample size (n)** to estimate a population proportion $p$ with $(1-\alpha)\times 100\%$ confidence level is:

$$
n=\frac{z_{\alpha/2}^2\times p^* (1-p^*)}{E^2} 
$$ {#eq-9.5}

**N.B:** From @eq-9.5 it can be proved that the sample size will be maximum for $p^*=0.5$. So to obtain the maximum sample size we can take $p^* (1-p^*)=0.5\times 0.5=0.25$ in @eq-9.5.

**Problem 9.13** [@anderson2020] At 95% confidence, how large a sample should be taken to obtain a margin of error of 0.03 for the estimation of a population proportion? Assume that past data are not available for developing a planning value for population proportion $p$ ?

**Problem 9.14** [@anderson2020] **Stay-at-Home Parenting.** In June 2014, Pew Research reported that in 16% of all homes with a stay-at-home parent, the father is the stay-at-home parent. An independent research firm has been charged with conducting a sample survey to obtain more current information.

a\. What sample size is needed if the research firm’s goal is to estimate the current proportion of homes with a stay-at-home parent in which the father is the stay-at-home parent with a margin of error of 0.03? Use a 95% confidence level.

b\. Repeat part (a) using a 99% confidence level.\

\

## Interval estimation: Comparing TWO populations
